---
title: "orthoRibbon Vignette: Ribbon-Diagram Visualization of Orthologous Genes Across Species"
author: "Jianhong Ou"
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('orthoRibbon')`"
abstract: >
  OrthoRibbon is a visualization toolkit for displaying orthologous
    gene relationships across species using ribbon-based diagrams.
    It enables intuitive comparison of conserved genes and genomic
    correspondence between chromosomes or genomic regions.
    The package supports flexible input formats and produces publication-quality
    figures for comparative genomics and evolutionary analysis..
vignette: >
  %\VignetteIndexEntry{orthoRibbon Vignette: Ribbon-Diagram Visualization of Orthologous Genes Across Species}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  html_document:
    theme: simplex
    toc: true
    toc_float: true
    toc_depth: 4
    fig_caption: true
---

```{r, echo=FALSE, results="hide", warning=FALSE}
knitr::opts_chunk$set(
    warning = FALSE, message = FALSE,
    fig.width = 12, fig.height = 4,
    dpi = 72
)
```
## Introduction

Comparative genomics plays a central role in understanding gene conservation and evolutionary relationships across species. Orthologous genes, which arise from speciation events and often retain similar biological functions, provide a fundamental framework for cross-species genomic comparison. Visualizing these relationships in an intuitive and scalable manner is important for interpreting genome organization and gene correspondence among multiple species.

OrthoRibbon is an R package designed to visualize orthologous gene relationships across species using ribbon-based diagrams. The package connects homologous genes between genomes through ribbons that link their genomic positions, enabling users to easily explore conserved genes and cross-species genomic structures. By transforming ortholog tables into clear visual representations, OrthoRibbon facilitates comparative genomic analysis and produces customizable, publication-quality figures.

OrthoRibbon supports flexible input formats and can integrate ortholog annotations from common databases such as Ensembl or other orthology inference tools. The package is designed to be lightweight and highly customizable, allowing users to control genomic layouts, ribbon aesthetics, and species ordering. These features make OrthoRibbon a practical visualization tool for researchers studying evolutionary conservation, genome organization, and cross-species gene relationships.


## Installation

```{r, eval=FALSE}
if (!require("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
BiocManager::install("jianhong/orthoRibbon")
```

## Quick start

There are five main steps to visualize your data with `orthoRibbon`:

Step 1. Load the required library.
```{r loadlibrary,warning = FALSE,message=FALSE}
library(orthoRibbon)
library(geneClusterPattern)
library(GenomeInfoDb)
```

Step 2. Define the species abbreviation. 
They are abbreviated species identifiers derived from the binomial (Latin) 
species names. You can try to guess it from common names.
The plot order is decided by the order of your input show as 'com_name' here. 

```{r getAbbreviations}
com_name <- geneClusterPattern::guessSpecies(
  c('human', 'house mouse', 'zebrafish'))
names(com_name) <- com_name # must assign the names.
com_name
```

Step 2. Retrieve chromosome, homologs, and gene coordinates via Biomart.

```{r retrieveHomologs,eval=FALSE}
# prepare mart for biomart
marts <- geneClusterPattern::guessSpecies(com_name, output='mart')

# prepare all the ensembl ids
ids <- orthoRibbon::getGeneIDs(com_name, marts)

# get chromosome info
chrom_infos<- lapply(com_name, GenomeInfoDb::getChromInfoFromEnsembl)

# retrieve all the homologs
homologs <- orthoRibbon::getHomologGRs(ids, com_name, marts)

# retrieve all the gene ranges
genes_gr <- orthoRibbon::getGeneGRs(ids, marts, homologs)

# reformat the homologs to data.frame
homolog_df <- getHomologIDs(homologs)
```

Since this step retrieve information via "biomaRt", it will take some time.
We will load presaved data.
```{r loadData}
extdata <- system.file('extdata', package='orthoRibbon')
chrom_infos <- readRDS(file.path(extdata, 'chrom_infos.rds'))
names(chrom_infos)
head(chrom_infos[[1]])
homologs_df <- readRDS(file.path(extdata, 'homolog_df.human.mouse.zebrafish.rds'))
head(homologs_df)
genes_gr <- readRDS(file.path(extdata, 'gene.GRanges.obj.rds'))
head(genes_gr)
```


Step 3. Prepare the plot data.

```{r prepareData}
set.seed(42)
plotData <- buildPlotData(com_name, homologs_df, genes_gr, chrom_infos,
                          chromosome_order_method = 'chr',
                          max_links=500 # small number for illustration
                          )
names(plotData)
```
You may want to try different methods to set the chromosome orders.
```{r otherOrderMethods,eval=FALSE}
# Different methods optimize different criteria:
# 1. GW (Gruvaeus-Wainer) - minimizes gradient measure
chr_order_gw <- buildPlotData(com_name, homologs_df, genes_gr, chrom_infos,
                          chromosome_order_method = "GW")

# 2. OLO (Optimal Leaf Ordering) - optimal for dendrograms
chr_order_olo <- buildPlotData(com_name, homologs_df, genes_gr, chrom_infos,
                          chromosome_order_method = "OLO")

# 3. TSP (Traveling Salesman Problem) - minimizes path length
chr_order_tsp <- buildPlotData(com_name, homologs_df, genes_gr, chrom_infos,
                          chromosome_order_method = "TSP")

# 4. Spectral - uses eigenvectors
chr_order_spec <- buildPlotData(com_name, homologs_df, genes_gr, chrom_infos,
                          chromosome_order_method = "Spectral")

# 5. customized
seq_names <- lapply(chrom_infos, function(.ele) sortSeqlevels(.ele$name))
chr_order_manu <- buildPlotData(com_name, homologs_df, genes_gr, chrom_infos,
                                chr_orders=seq_names)
```


Step 4. Calculate the exactly position for Bezier curve.

There are three methods to calculate the position, by chromosome locations of the gene, by the index of the gene, by the homolog chromosome names of the gene. 
```{r calBezierCurve}
# by coordinates
bezier_chr <- buildBezierDF(plotData$homolog_df_list,
                           colname1='topChr_finalOffset',
                           colname2='bottomChr_finalOffset')
# by the index
bezier_ix <- buildBezierDF(plotData$homolog_df_list,
                           colname1='topIx_finalOffset',
                           colname2='bottomIx_finalOffset')
# by homolog chromosome names
bezier_mini <- buildBezierDF(plotData$homolog_df_list,
                             colname1='topMini_finalOffset',
                             colname2='bottomMini_finalOffset')
```

Step 5. Plot the data.

```{r plotData}
# by coordinates
orthoRibbonPlot(com_name, bezier_chr,
                plotData$chrom_bars_df,
                plotData$chrom_label_df)
# by the index
orthoRibbonPlot(com_name, bezier_ix,
                plotData$chrom_bars_df,
                plotData$chrom_label_df)
# by homologous chromosome names
symbol_list_top <- plotData$symbol_list_top
symbol_list_top$x <- symbol_list_top$xMini
symbol_list_bottom <- plotData$symbol_list_bottom
symbol_list_bottom$x <- symbol_list_bottom$xMini
orthoRibbonPlot(com_name, bezier_mini,
                plotData$chrom_bars_df,
                plotData$chrom_label_df,
                show_symbol=TRUE,
                symbol_list_top=symbol_list_top,
                symbol_list_bottom=symbol_list_bottom,
                symbol_size=1)
```

## SessionInfo

```{r pressure, echo=FALSE}
sessionInfo()
```

